Chaotic mixing of tidal debris

### 2015-06-26 ms.tex
citations and references

% Context
%The large-scale, smooth components of galactic gravitational potentials are thought to be triaxial in shape. 
%Triaxial potentials that match observational constraints and measurements of the radial profile and axis ratios of dark matter halos in cosmological simulations are almost certainly not globally integrable. 
%An appreciable number of chaotic orbits are therefore expected in galactic halos.
%It is yet unclear whether chaotic regions of phase-space will affect the density evolution of stellar tidal streams, which form in the halos of galaxies from the disruption of stellar systems.
%% Aims
%Many numerical methods exist for assessing the degree of chaos for a given orbit (e.g., the Lyapunov exponent, the frequency diffusion rate). 
%These indicators, which are local to a single orbit, often predict a timescale over which chaos will be dynamically important.
%In this paper, we aim to compare these timescales to the density evolution of the finite-volume ensembles of orbits.
%% Methods: 
%We derp derp derp.
%% Results:
%We find that mean indicators such as the Lyapunov time or the (long term) frequency diffusion rate of a given ``parent'' orbit are not good predictors for the short-term density evolution of ensembles of orbits around this parent orbit.
%Instead, we find that the variance of the fundamental frequencies of the ensemble orbits over short timescales is a stronger indicator of the density evolution.
%% Conclusions:

@@

The dark matter halos of galaxies are thought to be triaxial in shape. However, despite suggestive evidence from a range of complimentary observational methods, this fundamental prediction from $\Lambda$CDM cosmology has not been conclusively verified. Around other galaxies, it is generally hard to measure the 3D mass profile because they are seen in projection. From the Earth's position within the Milky Way, our view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. The Milky Way halo has a low density of visible tracers, but luckily many of the halo stars are likely on non-random orbits and thus contain extra information (e.g., debris stripped from stellar systems).

@@

One example of stream modeling in a multi-component (static, analytic) potential was done by \citet{pearson15}, who tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} \citep{kuepper12} and N-body models. They use the observed SDSS number density of stream stars, a limited number of radial velocities for stream members, and the sky position, distance, radial velocity, and proper motion of the cluster itself to fit model streams to the data. In the oblate potential (a three-component bulge+disk+spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}: a three component bulge+disk+triaxial halo fit to Sagittarius stream data), the model streams generically form large, two-dimensional `fans' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness and curvature of the stream given the observational constraints of the present-day position and velocity of the cluster. The result in \citet{pearson15} demonstrates that the morphology of a stream alone can rule out a potential. With an understanding of the circumstances that lead to the differences in stream morphology, this could be a powerful tool for rejecting potentials from positional information alone.

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the triaxial potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further in this work.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position \citep[e.g.,][]{goldstein80}. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are generically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. With the inclusion of baryonic physics and sub-grid prescriptions for energy input due to supernovae and other feedback mechanisms, the inner potential ($\lesssim$$0.1R_{\rm vir}$ for a $\approx$$10^{12}~\msun$ halo mass) typically becomes more spherical, though the magnitude of this reshaping depends on the particular merger history and star formation efficiency within a given halo \citep[e.g.,][]{dubinski??, butsky15}. It is less clear what happens to the outer halo where tidal streams are more readily found. %; for this reason, we ask what we would expect to happen to tidal streams if the halo in which they form is strongly triaxial.

 \citet[][hereafter JS02]{jing02} found that a triaxial generalization of the NFW density profile \citep{navarro96} generates excellent fits to the density distributions within haloes in their high-resolution (dark-matter-only) N-body simulations, and they provide probability distributions for the axis ratios of a large sample of these halos. JS02 finds median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that JS02 use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are largely consistent with findings from more recent simulations \citep[e.g.,][]{millenium, hmm, butsky15} and consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. JS02 finds significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors \citep{romanowsky98}, radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11,butsky15}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.; see, e.g.,][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We generate an isoenergy grid of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass), $E$, chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits, which tend to plunge deep into the inner regions of the potential where the density profile is steep, however there are appreciable numbers of stochastic tube orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits which makes them more diffuse.

@@

\todo{apw}{motivate by saying we need a single measure to characterize ensemble evolution even though density evolution can be complex...can then map entire orbit grid}
\todo{apw}{These paragraphs need work, but still waiting for numerical results from the cluster...}
The density distributions do not change shape dramatically, even for the strongly chaotic ensemble; we therefore use the mean density as a metric for studying the density evolution of orbit ensembles. Figure~\ref{fig:densities} shows the mean density of the ensemble particles computed at 128 evenly-spaced intervals from the initial ensemble distribution to the distribution after 16 orbital periods. It is clear that the density of the regular orbit ensemble evolves slower than the weakly chaotic ensemble, and the strongly chaotic ensemble quickly becomes diffuse.

Motivated by the noticeable discrepancy between the regular and weakly chaotic orbit ensembles --- even though the chaotic timescale of the weakly chaotic ensemble parent orbit is thousands of orbital periods --- we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the mean density at the end of integration for each. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}), but now the greyscale indicates the final mean density. \todo{apw}{talk about the structure here, compare to freq. diff time map -- even though freq. diff time are long, can still resolve structure in these maps.}

@@

The potential considered in this work is `unrealistic' for the total potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{many}. However, the significance of baryonic relaxation or of time-dependence, triaxiality, and substructure on shaping the matter distribution within the Milky Way is largely unknown. Here we briefly summarize future directions for potential models:
\begin{itemize}
	\item \emph{Baryonic components}: \cite[][D08]{debattista08} and \cite[][V10]{valluri10} studied the orbit evolution induced by growing a baryonic disk in dark matter halos with various shapes and orientations (e.g., prolate, triaxial). In general, the authors find that the growth of a disk slowly deforms the orbits of mass tracers (e.g., dark matter particles) and preferentially populates tube orbits. Consequently, the inner shape of the potential becomes more oblate or spherical due to the prevalence of tube orbits. If the inner halos of galaxies are indeed close to spherical or oblate, the majority of orbits will be regular and chaos will be less important, however this is far from conclusive. We have found from simple experiments in superposing potential components that transition regions between potential shapes can actually enhance significant amounts of chaos. Though superposing potentials is not realistic, it is at least suggests that a more careful exploration of potential configurations is required to understand how complex, radius-dependent potential forms affect the amount and significance of chaos in galactic halos.

@@

\item \emph{Substructure}: Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep{??} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is very small \citep{??}. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

@@

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have numerically computed the magnitude of chaos for a large grid of iso-energy orbits in this potential using two independent methods that have been used extensively to classify and characterize chaotic orbits: the Lyapunov exponent and frequency diffusion rate. From each of these indicators, we compute a timescale over which chaos is likely to be important and find that the majority of orbits have chaotic timescales greater than 100s of orbital periods, however with the frequency diffusion rate we are still able to resolve weak chaos. We then study the density evolution of small ensembles of orbits generated around each orbit in the grid used for the previous experiment and find that along some orbits classified as weakly chaotic --- with chaotic timescales of 100s of orbital periods --- the orbit ensembles display enhanced density evolution and reach a lower overall density faster than orbit ensembles around nearby regular orbits (which mix due to phase-mixing alone). We explain this apparent discrepancy by considering the nature of chaotic diffusion: the classical chaos indicators are most sensitive to the slow, Arnold diffusion process that can cause large changes to orbital properties, but small-scale frequency evolution occurs over much shorter times as chaotic orbits stochastically diffuse across the stochastic layers that surround many resonances. We propose a new measure of chaos with tidal stream morphological evolution in mind: the amplitude of small-scale, fractional changes in frequencies over a given number of orbital periods along a stream progenitor orbit. When this amplitude is comparable to or larger than the typical spread in orbital properties of tidal debris from the progenitor, the phase-space density of the debris will evolve faster to a state of lower density relative to nearby regular orbits.

@@

 \todo{apw}{A sentence on broader implications / applications -- mapping the orbit structure of the potential by looking at streams}

@@

\appendix
\section{Lyapunov exponents} \label{sec:lyapapdx}
\todo{apw}{This section has not been edited carefully...}

%Using the definition of $w$ from Equation~\ref{eq:coords}, we can write Hamilton's equations as \footnote{A sum is implied with any repeated indices.}
%\begin{equation}
%	\dot{w}_i = \mathcal{J}_{ik}\,\frac{\partial H}{\partial w_k} \label{eq:ham}
%\end{equation}
%where $\mathcal{J}_{ik}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
%\begin{equation}
%	\mathcal{J}_{ik} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
%\end{equation}
%with $N$-dimensional identity matrices $\ident$. We will consider a nearby phase-space position, $w_i'$, separated from $w_i$ by an infinitesimal deviation, $\delta w_i$, such that $w_i' = w_i + \delta w_i$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as 
%\begin{align}
%	\dot{\delta w_i} &= \mathcal{J}_{ik} \, \frac{\partial^2 H}{\partial w_k \partial w_m} \, \delta w_m =  \mathcal{J}_{ik}\, D_{km} \, \delta w_m \\
%	&= A_{im} \, \delta w_m
%\end{align}
%where $D_{km}$ is the Hessian matrix evaluated at the parent orbit. The general solution for this equation is
%\begin{equation}
%	\delta w_i(t) = B_{im}(t) \, \delta w_i(0)
%\end{equation}
%where $\delta w_i(0)$ are the initial conditions for the deviation vector and $B_{im}$
% For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

>>>>

% Context
%The large-scale, smooth components of galactic gravitational potentials are thought to be triaxial in shape. 
%Triaxial potentials that match observational constraints and measurements of the radial profile and axis ratios of dark matter halos in cosmological simulations are almost certainly not globally integrable and therefore contain an appreciable number of chaotic orbits.
%However, estimates of the timescale over which chaos is likely to be important (e.g., the Lyapunov time) suggest that chaos may not be dynamically relevant over physically meaningful times.
% Aims:

% Methods:

%% Results:
% We find that indicators such as the Lyapunov time or the frequency diffusion time of an orbit are not good predictors for the short-time density evolution of ensembles of orbits (e.g., tidal debris) around this parent orbit.
% We propose a new measure for evaluating whether chaotic diffusion is important for tidal debris with a given spread in orbital properties over a given period of time: the variance of the fundamental frequencies of the ensemble orbits over short timescales is a stronger indicator of the density evolution.
% Conclusions:

@@

The dark matter halos of galaxies are thought to be triaxial in shape. However, despite suggestive evidence from a range of complimentary observational methods, this fundamental prediction from $\Lambda$CDM cosmology has not been conclusively verified. Around other galaxies, it is generally hard to measure the 3D mass profile because they are seen in projection. From the Earth's position within the Milky Way, our view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. The Milky Way halo has a low density of visible tracers, but luckily many of the halo stars are likely associated with debris stripped from stellar systems and thus contain extra information.

@@

One example of stream modeling in a multi-component (static, analytic) potential was done by \citet{pearson15}, who aimed to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} \citep{kuepper12} and N-body models. They used the observed SDSS number density of stream stars, a limited number of radial velocities for stream members, and the sky position, distance, radial velocity, and proper motion of the cluster itself to fit model streams to the data. In the oblate potential (a three-component bulge+disk+spherical halo potential), a thin model stream was easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}: a three component bulge+disk+triaxial halo fit to Sagittarius stream data), the model streams generically formed large, two-dimensional `fans' of debris near the ends, and no physically reasonable progenitor orbits could be found that reproduced the observed thinness and curvature of the stream given the observational constraints of the present-day position and velocity of the cluster. The result in \citet{pearson15} demonstrates that the morphology of a stream alone can rule out a potential. With an understanding of the circumstances that lead to the differences in stream morphology, this could be a powerful tool for rejecting potentials from positional information alone.

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the axisymmetric potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further in this work.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position \citep[e.g.,][]{goldstein80}. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are generically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. With the inclusion of baryonic physics and sub-grid prescriptions for energy input due to supernovae and other feedback mechanisms, the inner potential ($\lesssim$$0.1R_{\rm vir}$ for a $\approx$$10^{12}~\msun$ halo mass) typically becomes more spherical, though the magnitude of this reshaping depends on the particular merger history and star formation efficiency within a given halo \citep[e.g.,][though in Milky Way-like galaxies, baryonic disks will add non-sphericity to the total potential]{dubinski94,kazantzidis04, bryan13, butsky15}. It is less clear what happens to the outer halo.

 \citet[][hereafter JS02]{jing02} found that a triaxial generalization of the NFW density profile \citep{navarro96} generates excellent fits to the density distributions within haloes in their high-resolution (dark-matter-only) N-body simulations, and they provide probability distributions for the axis ratios of a large sample of these halos. JS02 find median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that JS02 use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are largely consistent with findings from more recent simulations \citep[e.g.,][]{bett07, veraciro11, butsky15} and consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. JS02 find significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

This potential is a simple (and unrealistic) model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors \citep{romanowsky98}, radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11,butsky15}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.; see, e.g.,][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We generate an isoenergy grid of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass), $E$, chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits, which tend to plunge deep into the inner regions of the potential where the density profile is steep, however there are appreciable numbers of stochastic tube orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes. Thin tidal streams may preferentially form along tube orbits rather than box orbits because of the faster disruption and debris diffusion expected for stellar systems on radially plunging orbits.

@@

The density evolution can be complex, however with a single measure that characterizes the ensemble density evolution we can map the entire grid of orbits used in previous sections (Figure~\ref{fig:icmap}). The density distributions do not change shape dramatically, even for the strongly chaotic ensemble. Figure~\ref{fig:densities} shows the evolution of the mean density of the ensemble particles computed at 256 evenly-spaced intervals from the initial ensemble distribution to the distribution after 16 orbital periods. It is clear that the density of the regular orbit ensemble evolves slower than the weakly chaotic ensemble, and the strongly chaotic ensemble quickly becomes diffuse; the mean density does not capture the full nature of the density evolution is apparently sensitive to the enhanced density evolution observed along the weakly chaotic orbit seen in Figure~\ref{fig:ensemble-density}.

Motivated by the noticeable discrepancy between the regular and weakly chaotic orbit ensembles --- even though the chaotic timescale of the weakly chaotic ensemble parent orbit is thousands of orbital periods --- we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the initial and final values of the mean, configuration-space density. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}, Figure~\ref{fig:icmap}), but now the greyscale indicates the ratio of the final mean density to the initial density. Much of the structure that is visible in the frequency diffusion time map (Figure~\ref{fig:freqdiff}) is again visible in this map of the density evolution of orbit ensembles. However, it is surprising that these features are present: the chaotic timescales predicted from both the Lyapunov and frequency diffusion times were typically 100s to 1000s of orbital periods for many of these resonance features.

@@

The potential considered in this work is `unrealistic' for the total potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{dubinski94,kazantzidis04, bryan13, butsky15}. However, the significance of baryonic relaxation or of time-dependence, triaxiality, and substructure on shaping the matter distribution within the Milky Way is largely unknown. Here we briefly summarize future directions for potential models:
\begin{itemize}
	\item \emph{Baryonic components}: \cite[][D08]{debattista08} and \cite[][V10]{valluri10} studied the orbit evolution induced by growing a baryonic disk in dark matter halos with various shapes and orientations (e.g., prolate, triaxial). In general, the authors find that the growth of a disk slowly deforms the orbits of mass tracers (e.g., dark matter particles) and preferentially populates tube orbits. Consequently, the inner shape of the potential becomes more oblate or spherical due to the prevalence of tube orbits. If the inner halos of galaxies are indeed close to spherical or oblate, the majority of orbits will be regular and chaos will be less important, however this is far from conclusive. We have found from simple experiments in superposing potential components that transition regions between potential shapes can actually enhance significant amounts of chaos. Though superposing potentials is not realistic, it is at least suggests that a more careful exploration of potential configurations is required to understand how complex, radius-dependent potential forms affect the amount and significance of chaos in galactic halos.

@@

\item \emph{Substructure}: Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep[e.g.,][]{diemand07} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is very small \citep{??}. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

@@

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have numerically computed the magnitude of chaos for a large grid of iso-energy orbits in this potential using two independent methods that have been used extensively to classify and characterize chaotic orbits: 1) the Lyapunov exponent and 2) the frequency diffusion rate. From each of these indicators, we compute a timescale over which chaos is likely to be important and find that the majority of orbits have chaotic timescales greater than 100s of orbital periods, however with the frequency diffusion rate we are still able to resolve weak chaos. We then study the density evolution of small ensembles of orbits generated around each orbit in the grid used for the previous experiment and find that along some orbits classified as weakly chaotic --- with chaotic timescales of 100s of orbital periods --- the orbit ensembles display enhanced density evolution and reach a lower overall density faster than orbit ensembles around nearby regular orbits (which mix due to phase-mixing alone). We explain this apparent discrepancy by considering the nature of chaotic diffusion: the classical chaos indicators are most sensitive to the slow, Arnold diffusion process that can cause large changes to orbital properties, but small-scale frequency evolution occurs over much shorter times as chaotic orbits stochastically diffuse across the stochastic layers that surround many resonances. We propose a new measure of chaos with tidal stream morphological evolution in mind: the amplitude of small-scale, fractional changes in frequencies over a given number of orbital periods along a stream progenitor orbit. When this amplitude is comparable to or larger than the typical spread in orbital properties of tidal debris from the progenitor, the phase-space density of the debris will evolve faster to a state of lower density relative to nearby regular orbits.

Our main conclusions are summarized as follows:
\begin{enumerate}
	\item the Lyapunov time and frequency diffusion time are powerful indicators of chaos, but do not capture the importance of small-scale chaotic diffusion for the density evolution of small ensembles of orbits (tidal debris);
	\item tidal debris becomes diffuse and thus harder to observe on weakly chaotic orbits when the small-scale chaotic diffusion of the fundamental frequencies has a scale comparable to the internal spread in frequencies of the debris;
	\item the details of the enhanced mixing along weakly chaotic orbits depends on the resonant structure of the potential;
	\item the variance of the fundamental frequencies of an orbit over a given time window is a strong predictor of the importance of small-scale chaotic frequency diffusion on the resulting morphology of tidal debris.
\end{enumerate}

@@

@@

\appendix
\section{Lyapunov exponents} \label{sec:lyapapdx}

%Using the definition of $w$ from Equation~\ref{eq:coords}, we can write Hamilton's equations as \footnote{A sum is implied with any repeated indices.}
%\begin{equation}
%	\dot{w}_i = \mathcal{J}_{ik}\,\frac{\partial H}{\partial w_k} \label{eq:ham}
%\end{equation}
%where $\mathcal{J}_{ik}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
%\begin{equation}
%	\mathcal{J}_{ik} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
%\end{equation}
%with $N$-dimensional identity matrices $\ident$. We will consider a nearby phase-space position, $w_i'$, separated from $w_i$ by an infinitesimal deviation, $\delta w_i$, such that $w_i' = w_i + \delta w_i$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as 
%\begin{align}
%	\dot{\delta w_i} &= \mathcal{J}_{ik} \, \frac{\partial^2 H}{\partial w_k \partial w_m} \, \delta w_m =  \mathcal{J}_{ik}\, D_{km} \, \delta w_m \\
%	&= A_{im} \, \delta w_m
%\end{align}
%where $D_{km}$ is the Hessian matrix evaluated at the parent orbit. The general solution for this equation is
%\begin{equation}
%	\delta w_i(t) = B_{im}(t) \, \delta w_i(0)
%\end{equation}
%where $\delta w_i(0)$ are the initial conditions for the deviation vector and $B_{im}$ is the solution matrix. 
% For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

If one is only interested in characterizing the degree of chaos, computing the full Lyapunov spectrum for an orbit is often not necessary. It is usually sufficient to compute an estimate of the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE). Using the definition of $\boldsymbol{w}$ from Equation~\ref{eq:coords}, consider an orbit that is a small deviation away from the parent orbit, $\boldsymbol{w}' = \boldsymbol{w} + \delta\boldsymbol{w}$. If the parent orbit is chaotic, the norm of the infinitesimal deviation should grow exponentially with time with some characteristic rate, $\lambda$,
\begin{equation}
	\|\delta\boldsymbol{w}(t)\| = e^{\lambda \, t} \, \|\delta\boldsymbol{w}_0\|
\end{equation}
\citep[see, e.g.,][]{lichtenberg83,tabor89}. From this expression, we see that
\begin{equation}
	\lambda(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|} \label{eq:mle}
\end{equation}
where the maximum Lyapunov exponent is the limit as $t\rightarrow \infty$,
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}\lambda(t). \label{eq:lmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to nonlinear evolution of the deviation and numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm \citep{benettin76}. There is no general way to determine $\tau$ except to perform convergence tests.
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	\lambda_N = \frac{1}{t_N}\sum_i^N \ln \frac{\|\delta \bs{w}(t_i)\|}{\|\delta \bs{w}_0\|} \label{eq:ftmle}
\end{equation}
where the $t_i$ are the times at which the renormalization occurs and the MLE is estimated after a very long time to approximate the limit of Equation~\ref{eq:lmax}.
